---
title: "Example: M&A Deal Announcements -- Power Analysis on Synthetic Data"
output: rmarkdown::html_vignette
bibliography: references.bib
---

```{r child="_setup.Rmd"}
```

> **SYNTHETIC DATA NOTICE.** Every observation in this example is generated by
> `simulate_event_study()` with a fixed seed. **No real M&A transactions or stock
> prices are used.** The example exists to teach event-study methodology and to
> demonstrate statistical *power* -- not to describe any actual deal.

## Abstract

Merger and acquisition announcements are a classic event-study application:
target shareholders typically earn large positive abnormal returns on the
announcement day. But how large must the true effect be, and how many deals must
you observe, before a test can *reliably* detect it? This example uses
[simulate_event_study()](../reference/simulate_event_study.html) to answer that
question directly, sweeping the true abnormal-return magnitude and reading off
the statistical power of the cross-sectional test.

For the underlying model and test, see
[Return Models](methods-return-models.html) and
[Test Statistics](methods-test-statistics.html).

## Research question

For a synthetic M&A deal sample, what minimum true abnormal return is needed for
the cross-sectional test to achieve roughly **80% power** at the 5% significance
level, holding the number of events fixed?

## Simulated data

`simulate_event_study()` draws firm returns from a market model with known
parameters, embeds a *known* event effect of a chosen magnitude, and runs the
event study `n_simulations` times, returning an `es_simulation` object. Its
components are the empirical `power`, a per-day `rejection_by_day` tibble, the
vector of realised `test_stats`, and the `params` used.

```{r simulate}
library(EventStudy)

sim_result <- simulate_event_study(
  n_events         = 15,
  event_window     = c(-5, 5),
  abnormal_return  = 0.01,   # true event-day effect embedded in the DGP
  n_simulations    = 200,
  seed             = 42
)

str(sim_result, max.level = 1)
cat("Empirical power at abnormal_return = 0.01:", sim_result$power, "\n")
```

## Power sweep

The primary result is a sweep over the true abnormal-return magnitude. For each
value we simulate afresh (fixed seed for reproducibility) and record the
empirical `power` -- the fraction of the `n_simulations` runs in which the test
correctly rejected the null on the event day.

```{r power-sweep}
# Extraction approach: simulate_event_study() returns an `es_simulation`
# object whose `$power` field is exactly the detection rate we want. We map
# over a grid of true effects and read $power from each run.
ar_grid <- c(0.001, 0.002, 0.004, 0.006, 0.010)

power_tbl <- purrr::map_dfr(ar_grid, function(ar) {
  sim <- simulate_event_study(
    n_events        = 15,
    event_window    = c(-5, 5),
    abnormal_return = ar,
    n_simulations   = 200,
    seed            = 42
  )
  tibble::tibble(true_abnormal_return = ar, power = sim$power)
})

knitr::kable(
  power_tbl, digits = 4,
  caption = "Detection power by true abnormal return (N = 15, cross-sectional test, alpha = 0.05)"
)
```

## Results plot

The `es_simulation` object has no built-in event-study plot (it is a simulation
summary, not a fitted task), so we render the power curve directly with
`plotly`. The dashed line marks the conventional 80% power target.

```{r power-plot, fig.width = 7, fig.height = 4}
plotly::plot_ly(
  power_tbl,
  x = ~true_abnormal_return, y = ~power,
  type = "scatter", mode = "lines+markers",
  line = list(color = "#20c997", width = 3),
  marker = list(color = "#20c997", size = 8)
) |>
  plotly::add_segments(
    x = min(ar_grid), xend = max(ar_grid), y = 0.8, yend = 0.8,
    line = list(color = "grey", dash = "dash"), showlegend = FALSE,
    inherit = FALSE
  ) |>
  plotly::layout(
    title = "Power curve: detection rate vs true abnormal return (synthetic)",
    xaxis = list(title = "True abnormal return"),
    yaxis = list(title = "Power", range = c(0, 1))
  )
```

## Power analysis interpretation

The power curve rises monotonically with the true effect size. At a true
abnormal return of 0.1% the test is essentially blind (power near the nominal
5% false-positive floor); by 1% the test detects the effect in the large
majority of samples. The practical reading: if a researcher expects M&A deal
announcements to move target prices by the customary several percent, a
sample of even fifteen deals gives ample power at conventional significance
levels. Weak or diffuse effects, by contrast, demand far larger samples --
the familiar power-vs-effect-size tradeoff [@MacKinlay1997].

## Design implications

Power is jointly determined by the true effect size, the number of events, the
event-window width, and the estimation-window length (which sharpens the
market-model fit and shrinks the abnormal-return standard error). Widening the
event window dilutes a concentrated announcement effect; lengthening the
estimation window tightens the benchmark. For a full treatment see the companion
[Monte Carlo Power Analysis](simulation-power-analysis.html) vignette and
[simulate_event_study()](../reference/simulate_event_study.html).

## Diagnostics note

On *real* (non-simulated) M&A samples, always check model-fit quality with
[es_diagnostics()](../reference/es_diagnostics.html): thin trading around small
targets and confounding news in the window are common threats that this
synthetic ideal-conditions example deliberately excludes.

## Further reading

- [Monte Carlo Power Analysis](simulation-power-analysis.html) -- full power-simulation vignette.
- [Test Statistics](methods-test-statistics.html) -- the cross-sectional test.
- [simulate_event_study()](../reference/simulate_event_study.html) -- reference page.
- [Bootstrap Inference & Multiple Testing](inference-robustness.html) -- robust alternatives.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
