---
title: "Example: Earnings Surprise Analysis (AAPL, MSFT, GOOGL)"
output: rmarkdown::html_vignette
bibliography: references.bib
---

```{r child="_setup.Rmd"}
```

## Abstract

This worked example takes you from the bundled `earnings_surprises` dataset --
three US mega-caps (AAPL, MSFT, GOOGL) around their Q1-2023 analyst-beat earnings
dates -- through a complete market-model event study: data loading, pipeline
execution, cross-sectional and standardized-residual test statistics (Patell Z
and BMP), and a rendered cumulative-abnormal-return (CAR) plot with written
interpretation.

For the theory behind the market model, see
[Return Models](methods-return-models.html); for the Patell Z and BMP tests, see
[Test Statistics](methods-test-statistics.html).

## Research question

Do AAPL, MSFT, and GOOGL earn significantly positive abnormal returns around
their quarterly earnings *beats*? Formally, we test the null

$$H_0:\ \mathbb{E}[AAR_t] = 0 \quad \text{for all } t \text{ in the event window},$$

against the alternative that the cumulative average abnormal return (CAAR)
differs from zero.

## Data

The `earnings_surprises` dataset ships with the package (in `data/`) and requires
**no network access**. It is a named list of three tibbles -- `firm` (stacked
daily adjusted prices for the three firms), `index` (the S&P 500 benchmark), and
`request` (the event specification: one row per firm with event date and
estimation/event windows) -- plus a `meta` provenance block.

```{r data}
library(EventStudy)
data("earnings_surprises", package = "EventStudy")

# The dataset is a list of tibbles; inspect the pieces we consume.
dplyr::glimpse(earnings_surprises$firm)
earnings_surprises$request[, c(
  "event_id", "firm_symbol", "index_symbol", "event_date",
  "group", "event_window_start", "event_window_end",
  "estimation_window_length"
)]
```

All three events belong to a single `"Earnings Beat"` group, which lets the
multi-event statistics aggregate them into one AAR/CAAR series.

## Setup and pipeline

`EventStudyTask$new()` takes the three tibbles positionally --
firm prices, reference (index) prices, and the request table. We fit a
`MarketModel` and attach the multi-event statistics we care about
(`CSectTTest`, `PatellZTest`, `BMPTest`) via a `MultiEventStatisticsSet`.

```{r pipeline}
task <- EventStudyTask$new(
  earnings_surprises$firm,    # firm daily adjusted prices
  earnings_surprises$index,   # benchmark index prices
  earnings_surprises$request  # event / window specification
)

params <- ParameterSet$new(
  return_model = MarketModel$new(),
  multi_event_statistics = MultiEventStatisticsSet$new(
    tests = list(CSectTTest$new(), PatellZTest$new(), BMPTest$new())
  )
)

result <- run_event_study(task, params)
```

The pipeline builds the task, fits the market model on each firm's estimation
window, computes abnormal returns over the event window, and evaluates the
statistics. See [run_event_study()](../reference/run_event_study.html),
[EventStudyTask](../reference/EventStudyTask.html),
[ParameterSet](../reference/ParameterSet.html),
[MarketModel](../reference/MarketModel.html),
[PatellZTest](../reference/PatellZTest.html), and
[BMPTest](../reference/BMPTest.html).

## Results table

The `tidy()` method returns per-firm cumulative abnormal returns as a tidy
tibble. We call the S3 method explicitly (`tidy.EventStudyTask()`), since the
`broom` generic is not re-exported by the package.

```{r car-table}
car_tbl <- EventStudy::tidy.EventStudyTask(result, type = "car")

# Show the last (widest) CAR window per firm.
car_tbl |>
  dplyr::group_by(firm_symbol) |>
  dplyr::slice_tail(n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(firm_symbol, term, estimate, std.error, statistic, p.value) |>
  knitr::kable(
    digits  = 4,
    caption = "Cumulative Abnormal Returns by firm (widest event window)"
  )
```

## Plot

`plot_event_study()` returns a `ggplot`; wrapping it with `plotly::ggplotly()`
produces an interactive plot for the article.

```{r car-plot, fig.width = 7, fig.height = 4}
gg <- plot_event_study(
  result,
  type    = "car",
  event_id = 1L,
  title   = "AAPL: cumulative abnormal return with 95% confidence band"
)
plotly::ggplotly(gg)
```

## Statistical interpretation

The Patell Z and BMP statistics both aggregate the three firms' standardized
abnormal returns, but BMP additionally corrects for event-induced
cross-sectional variance.

```{r stat-values}
pz  <- result$aar_caar_tbl$PatellZ[[1]]
bmp <- result$aar_caar_tbl$BMP[[1]]

caar_patell <- round(utils::tail(pz$caar,    1L), 4)
z_patell    <- round(utils::tail(pz$caar_z,  1L), 3)
t_bmp       <- round(utils::tail(bmp$cbmp_t, 1L), 3)
df_bmp      <- utils::tail(bmp$n_valid_events, 1L) - 1L

# Two-sided p-values: Patell Z ~ N(0,1); BMP t ~ t(df_bmp)
p_patell <- round(2 * pnorm(abs(z_patell), lower.tail = FALSE), 4)
p_bmp    <- round(2 * pt(abs(t_bmp), df = df_bmp, lower.tail = FALSE), 4)
```

At the end of the event window the CAAR across the three firms is
`r caar_patell`. The Patell Z on the cumulative series is `r z_patell`
(two-sided p = `r p_patell`) and the BMP statistic is `r t_bmp`
(two-sided p = `r p_bmp`). Where the two disagree in magnitude, the BMP figure
is the more conservative -- it inflates the standard error when abnormal returns
are correlated across firms on the event day, which is exactly the situation a
common earnings-season calendar can create [@BMP1991]. Read together,
they tell you whether an apparently significant AAR survives a cross-sectional
correlation correction.

## Diagnostics note

The market-model OLS residuals underpinning these statistics can be tested for
autocorrelation and non-normality with
[es_diagnostics()](../reference/es_diagnostics.html); a poor estimation-window
fit weakens every downstream statistic. See
[Diagnostics & Robustness](methods-diagnostics.html) for the full battery.

## Further reading

- [Return Models](methods-return-models.html) -- market-model theory and alternatives.
- [Test Statistics](methods-test-statistics.html) -- Patell Z and BMP derivations.
- [run_event_study()](../reference/run_event_study.html) -- pipeline reference page.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
