---
title: "Example: Regulatory Shock -- Dieselgate VW-Group vs Peers"
output: rmarkdown::html_vignette
bibliography: references.bib
---

```{r child="_setup.Rmd"}
```

## Abstract

This worked example uses the bundled `dieselgate` dataset to demonstrate a
**two-group** event study that isolates an idiosyncratic shock from an
industry-wide one. Volkswagen and Porsche (the directly-exposed *VW Group*) are
compared against peer automakers BMW and Mercedes-Benz (*Other*). By contrasting
the cumulative average abnormal return (CAAR) across groups around the
September 2015 emissions disclosure, we ask whether the scandal caused abnormal
losses *beyond* the German auto sector as a whole.

For the market model, see [Return Models](methods-return-models.html); for the
cross-sectional and sign tests, see [Test Statistics](methods-test-statistics.html).

## Research question

Did the VW emissions announcement cause significantly more negative cumulative
abnormal returns for the VW corporate family (VOW.DE, PAH3.DE) than for the
industry-peer group (BMW.DE, MBG.DE)? We test, **separately for each group**,

$$H_0:\ \mathbb{E}[CAAR] = 0,$$

and then compare the two groups' CAR distributions directly.

## Data

The `dieselgate` dataset ships in `data/` and needs **no network access**. Like
the other bundled datasets it is a named list -- `firm` (all four automakers'
daily adjusted prices, stacked), `index` (the benchmark), a four-row `request`
carrying the per-firm event windows and the crucial `group` assignment, and a
`meta` provenance block. See the provenance notes in `?dieselgate`.

```{r data}
library(EventStudy)
data("dieselgate", package = "EventStudy")

dplyr::glimpse(dieselgate$firm)

# The group column drives the two-group comparison.
dplyr::count(dieselgate$request, group)
```

## Setup and pipeline

We build the task from the three tibbles and fit a market model, attaching the
cross-sectional t-test (`CSectTTest`) and the `SignTest` as multi-event
statistics. Both aggregate the firm-level abnormal returns *within each group*.

```{r pipeline}
task <- EventStudyTask$new(
  dieselgate$firm,
  dieselgate$index,
  dieselgate$request
)

params <- ParameterSet$new(
  return_model = MarketModel$new(),
  multi_event_statistics = MultiEventStatisticsSet$new(
    tests = list(CSectTTest$new(), SignTest$new())
  )
)

result <- run_event_study(task, params)
```

See [ParameterSet](../reference/ParameterSet.html),
[MultiEventStatisticsSet](../reference/MultiEventStatisticsSet.html),
[CSectTTest](../reference/CSectTTest.html), and
[SignTest](../reference/SignTest.html).

## Results table

The per-group CAAR at the end of the event window is the headline result. We
pull it from the tidy AAR/CAAR output, and summarise the raw per-firm CARs by
group with [car_by_group()](../reference/car_by_group.html).

```{r caar-table}
caar_tbl <- EventStudy::tidy.EventStudyTask(result, type = "aar")

caar_tbl |>
  dplyr::group_by(group) |>
  dplyr::slice_tail(n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(group, term, caar, caar_statistic, caar_p.value) |>
  es_tt(
    digits  = 4,
    caption = "CAAR by group at the end of the event window: VW-Group vs Peers"
  )

# Group-level CAR summary and a between-group difference test.
grp <- car_by_group(result)
es_tt(grp$summary, digits = 4,
      caption = "Per-firm CAR summary by group")
```

## Plot

```{r caar-plot, fig.width = 7, fig.height = 4}
gg <- plot_event_study(
  result,
  type  = "aar",
  group = "VW Group",
  title = "VW Group: cumulative average abnormal return with 95% band"
)
plotly::ggplotly(gg)
```

## Interpretation

```{r stat-values}
vw_caar  <- result$aar_caar_tbl$CSectT[[
  which(result$aar_caar_tbl$group == "VW Group")]]
oth_caar <- result$aar_caar_tbl$CSectT[[
  which(result$aar_caar_tbl$group == "Other")]]

vw_end   <- round(utils::tail(vw_caar$caar,   1L), 4)
vw_t     <- round(utils::tail(vw_caar$caar_t, 1L), 2)
oth_end  <- round(utils::tail(oth_caar$caar,  1L), 4)
oth_t    <- round(utils::tail(oth_caar$caar_t, 1L), 2)
```

By the end of the window the *VW Group* CAAR is `r vw_end`
(cross-sectional *t* = `r vw_t`), while the *Other* group's CAAR is `r oth_end`
(*t* = `r oth_t`). The contrast is the whole story: VW Group firms crater while
peer automakers barely move. Because the peer group -- exposed to the same
macro and sector conditions -- shows no comparable drop, the shock is
**idiosyncratic** to the VW corporate family rather than an industry-wide
contagion. The `SignTest` corroborates the direction non-parametrically: with
every VW-group firm posting negative CARs, the sign statistic points the same
way as the parametric test, which matters when a two-firm group makes the
normality assumption of the *t*-test fragile [@Brown1985]. This two-group
design is the standard technique for separating firm-specific from sector-wide
regulatory shocks.

## Diagnostics note

With only a handful of firms per group, groupwise residual autocorrelation and
non-normality materially affect the standard errors. Check the estimation-window
fit with [es_diagnostics()](../reference/es_diagnostics.html) before trusting the
parametric *p*-values.

## Further reading

- [Return Models](methods-return-models.html) -- market-model theory.
- [Test Statistics](methods-test-statistics.html) -- cross-sectional and sign tests.
- [Panel / DiD](methods-panel-did.html) -- for staggered-treatment regulatory events.
- [car_by_group()](../reference/car_by_group.html) -- group CAR summary reference.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
