---
title: "Methods: Synthetic Control"
output: rmarkdown::html_vignette
bibliography: references.bib
# pkgdown article — excluded from the tarball via .Rbuildignore (^vignettes/articles).
---

```{r child="_setup.Rmd"}
```

## 1. Title & Abstract

This article covers the **synthetic control method (SCM)** for estimating the
effect of a single treated unit against a data-driven weighted combination of
untreated "donor" units. SCM is the tool of choice when one unit is treated,
many candidate controls exist, and no single control is a credible
counterfactual on its own. It is estimated live here on inline synthetic
treated-plus-donor data using the base-R `method = "optim"` solver.

## 2. When to Use This Method

Use SCM for **comparative case studies**: one region, firm, or market receives
a treatment (a law, a shock, an entry) and you need a counterfactual. Rather
than pick one comparison unit, SCM forms a convex weighted average of donors
that tracks the treated unit's **pre-treatment** trajectory, then reads the
post-treatment gap as the effect [@AbadieDiamondHainmueller2010].

## 3. Intuition

Blend the donor pool so the blend's pre-treatment path lies on top of the
treated unit's path. If the synthetic unit and the real unit were
indistinguishable before treatment, their divergence afterwards is the
treatment effect. The blend weights are non-negative and sum to one, keeping
the synthetic control inside the donors' support (no extrapolation).

## 4. Model & Null Hypothesis

The synthetic counterfactual is a convex combination of donor outcomes, with
weights chosen to minimise pre-period mean squared prediction error (MSPE):

<!-- Formula verified: R/synthetic_control.R:183,226 matches AbadieDiamondHainmueller2010 -->

$$
\hat{Y}^{N}_{1t} = \sum_{j=2}^{J+1} w_j\, Y_{jt},
\qquad
w_j \geq 0,\ \sum_j w_j = 1,
$$

and the treatment effect is the post-treatment gap:

$$
\text{gap}_t = Y_{1t} - \hat{Y}^{N}_{1t}.
$$

The null hypothesis is $H_0: \text{gap}_t = 0$ for $t \geq t_0$ — the treated
unit tracks its synthetic counterfactual after treatment, i.e. no effect.

## 5. Assumptions

- **Good pre-treatment fit:** a low pre-period MSPE; a poor fit invalidates the
  counterfactual.
- **Convex hull:** the treated unit's pre-path lies within the donors' span
  (weights on the simplex, no extrapolation).
- **No interference / spillovers:** donors are unaffected by the treatment.
- **No anticipation** before $t_0$.

## 6. Worked Example

We build inline synthetic data in base R with an explicit local `set.seed(42)`:
5 donors over 80 periods (long format `unit, time, outcome`) and a treated unit
(`time, outcome`) that tracks the donor average pre-treatment and jumps after
$t_0 = 60$. We default to `method = "optim"` (pure `stats::optim` L-BFGS-B, no
optional package).

```{r build-sc}
library(EventStudy)
set.seed(42)

periods <- 1:80
ttime   <- 60
donor_ids <- paste0("D", 1:5)

donor_data <- do.call(rbind, lapply(donor_ids, function(d) {
  data.frame(unit = d, time = periods, outcome = cumsum(rnorm(80, 0.1, 1)))
}))

# Treated = donor-average pre-treatment + a post-treatment jump; lengths match 80.
donor_mean <- as.numeric(tapply(donor_data$outcome, donor_data$time, mean))
post_jump  <- ifelse(periods >= ttime, 0.15 * (periods - ttime + 1), 0)
treated_data <- data.frame(
  time    = periods,
  outcome = donor_mean + post_jump + rnorm(80, 0, 0.3)
)

task <- SyntheticControlTask$new(treated_data, donor_data, treatment_time = ttime)
res  <- estimate_synthetic_control(task, method = "optim")   # base R stats::optim
```

The `quadprog` quadratic-programming solver is available when that optional
package is installed:

```{r sc-quadprog, eval=requireNamespace("quadprog", quietly=TRUE)}
res_qp <- estimate_synthetic_control(task, method = "quadprog")
```

## 7. Rendered Table

```{r results-table}
knitr::kable(
  head(res$results$trajectory, 10),
  caption = "Treated vs synthetic trajectory with the estimated gap (first 10 periods)."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Gap between the treated unit and its synthetic control over time."}
traj <- res$results$trajectory
plot(traj$time, traj$gap, type = "l",
     xlab = "Time", ylab = "Gap (treated - synthetic)",
     main = "Synthetic control gap")
abline(v = 60, lty = 2)
abline(h = 0, col = "grey60")
```

## 9. Interpretation

Before $t_0 = 60$ the gap should hover near zero — the synthetic control tracks
the treated unit, confirming a credible counterfactual (low pre-period MSPE in
`res$results$pre_mspe`). After $t_0$ the gap opens up, recovering the built-in
post-treatment jump; its average is the ATT in `res$results$att`. A gap that
was already non-zero *before* treatment would signal a poor fit and an
untrustworthy estimate. Inference typically compares this treated-unit gap to
the distribution of placebo gaps obtained by reassigning treatment to each
donor.

## References
