---
title: "Methods: Intraday Event Studies"
output: rmarkdown::html_vignette
bibliography: references.bib
# pkgdown article — excluded from the tarball via .Rbuildignore (^vignettes/articles).
---

```{r child="_setup.Rmd"}
```

## 1. Title & Abstract

This article covers **intraday event studies** — abnormal-return analysis at
minute (or finer) frequency rather than daily. Many events (earnings releases,
regulatory announcements, central-bank communications) resolve within minutes,
and daily data averages that signal away. EventStudy's `IntradayEventStudyTask`
runs the same market-model machinery on POSIXct-timestamped high-frequency
bars. Here it is estimated live on an inline synthetic one-minute session.

## 2. When to Use This Method

Use an intraday study when the event's effect is expected to materialise and
possibly decay **within a trading day**, and you have timestamped bar or tick
data. The frequency lets you locate the exact minute of price impact and
measure how quickly the market absorbs the news [@BarclayWarner1993] — detail
that a daily study cannot resolve.

## 3. Intuition

The idea is identical to the daily market model, only the clock ticks faster.
Fit the firm's minute returns on the market's minute returns over a pre-event
estimation window, then measure minute-by-minute abnormal returns around the
event timestamp. Cumulating them across the intraday event window traces the
price-impact path in real time.

## 4. Model & Null Hypothesis

At intraday frequency $\tau$, the abnormal return is the market-model residual
computed by the same shared engine used for daily studies:

<!-- Formula verified: R/models.R:248 (shared AR engine) matches BarclayWarner1993 -->

$$
AR_{i,\tau} = r_{i,\tau} - \hat{\alpha}_i - \hat{\beta}_i\, r_{m,\tau}.
$$

The null hypothesis is $H_0: \mathbb{E}[AR_{i,\tau}] = 0$ across the intraday
event window — no abnormal minute-level return around the event.

## 5. Assumptions

- **Timestamp alignment:** firm and reference bars share a common time grid;
  timestamps must be `POSIXct`.
- **Stable intraday beta** estimated in the pre-event window.
- **Microstructure caveats:** bid-ask bounce and thin trading inflate
  high-frequency noise; keep the bar size coarse enough to be meaningful.
- **Clean estimation window** free of overlapping intraday events.

## 6. Worked Example

We construct an inline synthetic intraday dataset in base R / `tibble` with an
explicit local `set.seed(42)` for byte-stable output. Column names match the
existing `intraday-event-study` vignette: firm and reference tables carry
`symbol`, `timestamp` (POSIXct), `price`; the request carries `event_id`,
`group`, `firm_symbol`, `index_symbol`, `event_timestamp`, and minute windows.

```{r build-intraday}
library(EventStudy)
library(tibble)
set.seed(42)

ts <- seq(as.POSIXct("2023-06-01 09:30:00", tz = "UTC"),
          by = 60, length.out = 300)   # 300 one-minute bars

firm_data <- tibble(
  symbol    = "FIRM_A",
  timestamp = ts,                       # MUST be POSIXct or the constructor stops
  price     = 100 * cumprod(1 + rnorm(300, 0, 0.001))
)
index_data <- tibble(
  symbol    = "INDEX_1",
  timestamp = ts,
  price     = 1000 * cumprod(1 + rnorm(300, 0, 0.0008))
)
request <- tibble(
  event_id                 = 1L,
  group                    = "Intraday",
  firm_symbol              = "FIRM_A",
  index_symbol             = "INDEX_1",
  event_timestamp          = as.POSIXct("2023-06-01 12:30:00", tz = "UTC"),
  event_window_start       = -30L,
  event_window_end         = 30L,
  shift_estimation_window  = -31L,
  estimation_window_length = 120L
)

task   <- IntradayEventStudyTask$new(firm_data, index_data, request)
params <- ParameterSet$new()
task   <- prepare_intraday_event_study(task, params)
task   <- fit_model(task, params)
task   <- calculate_statistics(task, params)
```

## 7. Rendered Table

```{r results-table}
es_tt(
  head(tidy.EventStudyTask(task, type = "car"), 10),
  caption = "Intraday cumulative abnormal returns around the event minute (first 10 offsets)."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Intraday abnormal-return path around the event timestamp."}
plot_event_study(task)
```

## 9. Interpretation

The table in §7 reports minute-level cumulative abnormal returns relative to the
event timestamp; the corresponding $t$-statistics flag which minutes carry a
significant response. The plot in §8 shows the intraday path — a sharp break at
offset zero followed by a plateau is the signature of near-instant price
discovery, whereas a gradual drift suggests slower information diffusion. Because
the underlying abnormal-return engine (§4) is shared with the daily pipeline,
every downstream statistic and plot behaves identically; only the clock changes.

## References
