---
title: "Methods: Panel / Difference-in-Differences"
output: rmarkdown::html_vignette
bibliography: references.bib
# pkgdown article — excluded from the tarball via .Rbuildignore (^vignettes/articles).
---

```{r child="_setup.Rmd"}
```

## 1. Title & Abstract

This article covers EventStudy's **panel / difference-in-differences (DiD)**
estimators for staggered-adoption designs, where units are treated at
different times. It contrasts the classic two-way fixed-effects (TWFE)
estimator with the modern heterogeneity-robust estimators (Sun-Abraham,
Callaway-Sant'Anna, de Chaisemartin-D'Haultfoeuille, Borusyak-Jaravel-Spiess),
and renders a dynamic event-study estimate live on an inline synthetic panel.

## 2. When to Use This Method

Use a panel event study when treatment is absorbing and **staggered** across
units, and you want event-time (leads/lags) dynamics rather than a single
average effect. Reach for the modern estimators when treatment effects are
**heterogeneous across cohorts or time**, because standard TWFE can then be
biased by "forbidden comparisons" that use already-treated units as controls
[@GoodmanBacon2021].

## 3. Intuition

Static TWFE nets out unit and time averages and reads the residual jump at
treatment as the effect. Dynamic TWFE traces that jump across event time,
using the period just before treatment ($k = -1$) as the baseline. The modern
estimators repair TWFE by only ever comparing treated units to *not-yet-treated
or never-treated* units, cohort by cohort.

## 4. Model & Null Hypothesis

**Static TWFE** with unit effects $\alpha_i$, time effects $\lambda_t$, and a
single treatment dummy $D_{it}$:

<!-- Formula verified: R/panel_event_study.R:210-214 matches GoodmanBacon2021 -->

$$
y_{it} = \alpha_i + \lambda_t + \delta D_{it} + \varepsilon_{it}.
$$

**Dynamic / event-study TWFE** replaces the single dummy with event-time
indicators relative to each unit's treatment time $g_i$, omitting the base
period $k = -1$:

<!-- Formula verified: R/panel_event_study.R:242,126 matches SunAbraham2021 -->

$$
y_{it} = \alpha_i + \lambda_t
       + \sum_{k \neq -1} \beta_k \,\mathbf{1}\{t - g_i = k\}
       + \varepsilon_{it}.
$$

**Sun-Abraham** estimates cohort-by-relative-time interaction effects and
aggregates them with cohort-share weights, avoiding the contamination that
biases plain dynamic TWFE under heterogeneity:

<!-- Formula verified: R/panel_event_study.R:316 matches SunAbraham2021 -->

The null hypothesis is $H_0: \beta_k = 0$ for all $k \geq 0$ — no dynamic
treatment effect at any post-treatment horizon.

## 5. Assumptions

- **Parallel trends** (conditional on fixed effects) between treated and
  comparison units.
- **No anticipation** before treatment time $g_i$.
- **Absorbing treatment:** once treated, always treated.
- Modern estimators additionally allow **heterogeneous** effects across cohorts
  and relative time; TWFE assumes homogeneity for an unbiased average.

## 6. Worked Example

We build an inline synthetic staggered panel in base R with an explicit local
`set.seed(42)` (on top of the `_setup.Rmd` seed) so the rendered coefficients
are byte-stable across rebuilds: 10 units over 10 periods, 3 staggered cohorts.

```{r build-panel}
library(EventStudy)
set.seed(42)

n_units <- 10
periods <- 1:10
cohorts <- c(4, 6, 8)   # 3 staggered treatment cohorts

panel <- do.call(rbind, lapply(1:n_units, function(u) {
  g <- cohorts[((u - 1) %% 3) + 1]
  data.frame(
    unit_id        = u,
    time_id        = periods,
    treatment_time = g,
    treated        = as.integer(periods >= g),
    y              = 0.5 * u + 0.3 * periods +
                     1.5 * (periods >= g) +
                     rnorm(length(periods), 0, 0.5)
  )
}))

task <- PanelEventStudyTask$new(
  panel,
  unit_id        = "unit_id",
  time_id        = "time_id",
  outcome        = "y",           # passed explicitly, not relying on the default
  treatment      = "treated",
  treatment_time = "treatment_time"
)
```

The **live** estimators use only `stats::lm()` and are always available. The
entry point is the function `estimate_panel_event_study(task, method = ...)`:

```{r estimate-live}
res <- estimate_panel_event_study(task, method = "dynamic_twfe", leads = 3, lags = 3)

# static_twfe and sun_abraham are equally live (base R):
res_static <- estimate_panel_event_study(task, method = "static_twfe")
res_sa      <- estimate_panel_event_study(task, method = "sun_abraham", leads = 3, lags = 3)
```

The **modern external-package** estimators are shown conceptually only — they
require optional packages and are gated `eval=FALSE`. Note the **exact** source
method strings (not the shorthand):

```{r estimate-optional, eval=FALSE}
estimate_panel_event_study(task, method = "callaway_santanna")             # needs 'did'
estimate_panel_event_study(task, method = "dechaisemartin_dhaultfoeuille")  # needs 'DIDmultiplegt'
estimate_panel_event_study(task, method = "borusyak_jaravel_spiess")        # needs 'didimputation'
```

## 7. Rendered Table

```{r results-table}
knitr::kable(
  res$results$coefficients,
  caption = "Dynamic TWFE event-time coefficients (base period k = -1) on the synthetic staggered panel."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Dynamic TWFE event-study coefficients across relative time."}
plot_panel_event_study(res)
```

## 9. Interpretation

The pre-treatment coefficients (relative time $< 0$) should hover near zero — a
visual parallel-trends check. The post-treatment coefficients recover the
built-in effect (a level shift of $1.5$ at and after treatment). If the pre
coefficients trended, the parallel-trends assumption in §5 would be suspect and
the estimate untrustworthy. When cohort effects differ, prefer Sun-Abraham or
the external estimators — plain dynamic TWFE can attribute one cohort's dynamics
to another through forbidden comparisons [@CallawaySantAnna2021;
@SunAbraham2021; @deChaisemartinDHaultfoeuille2020; @BorusyakJaravelSpiess2024].

## References
