---
title: "Methods: Return Models"
output: rmarkdown::html_vignette
bibliography: references.bib
# NOTE: No vignette: block — this is a pkgdown article, excluded from the
# tarball via .Rbuildignore (^vignettes/articles). Not a CRAN vignette.
---

```{r child="_setup.Rmd"}
```

## 1. Title & Abstract

This article explains the **return models** that EventStudy uses to generate
abnormal returns — the raw material of every event study. A return model
specifies what a security's return *would have been* absent the event; the
abnormal return is the gap between the realised return and that expectation.
After reading, you will know which model family fits which research setting,
see the market model estimated live on the bundled `dieselgate` data, and know
where the factor-based (Fama-French / Carhart), conditional-volatility (GARCH),
and buy-and-hold (BHAR) variants apply.

## 2. When to Use This Method

The choice of return model is driven by data availability and the return
horizon:

- **Market model** — the workhorse. Needs only a firm return series and one
  market index. Appropriate for short-horizon daily event studies.
- **Market-adjusted / comparison-period-mean** — when you cannot estimate a
  reliable beta (short history), constrain $\beta=1$ or use the pre-event mean.
- **Fama-French 3/5-factor, Carhart 4-factor** — when systematic exposure to
  size, value, profitability, investment, or momentum could contaminate a
  simple market beta. Require a factor return table.
- **GARCH / DCC-GARCH** — when event-induced or clustered volatility makes the
  homoskedastic OLS residual variance untrustworthy.
- **BHAR (buy-and-hold abnormal returns)** — long-horizon studies where
  compounding, not summing, is the economically correct aggregation.

## 3. Intuition

Fit a line through the firm's returns versus the market's returns during a
quiet **estimation window**. That line is the firm's "normal" behaviour. Carry
the line forward into the **event window** and measure how far each realised
return sits above or below it. Those vertical distances are the abnormal
returns; their running total is the cumulative abnormal return (CAR).

## 4. Model & Null Hypothesis

The **market model** regresses firm returns on contemporaneous market returns
over the estimation window:

<!-- Formula verified: R/models.R:248 matches MacKinlay1997 eq. 1-2 -->

$$
R_{it} = \alpha_i + \beta_i R_{mt} + \varepsilon_{it},
\qquad
\hat{\varepsilon}_{it} = R_{it} - (\hat{\alpha}_i + \hat{\beta}_i R_{mt}).
$$

The abnormal return $\hat{\varepsilon}_{it}$ is the out-of-sample residual in
the event window. Cumulating over $[t_1, t_2]$ gives the CAR:

<!-- Formula verified: R/single_event_test_statistics.R:127 matches MacKinlay1997 eq. 15 -->

$$
\text{CAR}(t_1, t_2) = \sum_{t=t_1}^{t_2} \hat{\varepsilon}_{t}.
$$

For long horizons, buy-and-hold abnormal returns compound rather than sum:

<!-- Formula verified: R/single_event_test_statistics.R:189-191 matches BarberLyon1997 -->

$$
\text{BHAR}_i = \prod_{t}(1 + R_{it}) - \prod_{t}(1 + R_{mt}).
$$

The null hypothesis is $H_0: \mathbb{E}[\hat{\varepsilon}_{it}] = 0$ in the
event window — the event conveys no abnormal information [@MacKinlay1997;
@Brown1985].

## 5. Assumptions

- **Return-generating stability:** $\alpha_i, \beta_i$ estimated in the
  estimation window carry into the event window (testable via rolling betas).
- **Uncorrelated, homoskedastic residuals** for OLS inference (relaxed by the
  GARCH family; see §8).
- **Clean estimation window:** no confounding events contaminate the pre-event
  data used to fit the model.
- **Correct market proxy:** the index spans the firm's systematic risk
  (relaxed by the factor models).

## 6. Worked Example

The always-live path here is the **MarketModel** family on the bundled
`dieselgate` data — no factor table or optional package required. Models are
selected by passing an **R6 model object** into `ParameterSet$new()`, never a
string.

```{r load-data}
library(EventStudy)
data("dieselgate")
```

```{r fit-pipeline}
task <- EventStudyTask$new(
  firm_stock_data_tbl = dieselgate$firm,
  reference_tbl       = dieselgate$index,
  request_tbl         = dieselgate$request
)

# Model selection is by R6 object, not a string:
params <- ParameterSet$new(return_model = MarketModel$new())

task <- prepare_event_study(task, params)
task <- fit_model(task, params)
task <- calculate_statistics(task, params)
```

`MarketAdjustedModel$new()`, `ComparisonPeriodMeanAdjustedModel$new()`,
`RollingWindowModel$new()`, and `BHARModel$new()` are also fully live on
`dieselgate` — swap the `return_model` argument to use them.

## 7. Rendered Table

```{r results-table}
es_tt(
  head(tidy.EventStudyTask(task, type = "car"), 10),
  caption = "Market-model CAR t-statistics on the dieselgate event (first 10 rows)."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Cumulative abnormal return over the dieselgate event window (event 1)."}
plot_event_study(task, type = "car", event_id = 1)
```

## 8b. Factor and Conditional-Volatility Variants

**Fama-French 3/5-factor and Carhart 4-factor** models require a factor table
with `market_excess`, `smb`, `hml` (plus `rmw`, `cma` for FF5 and `mom` for
Carhart). The bundled `dieselgate` data has **no factor table**, so the
following is a conceptual snippet (`eval=FALSE`) — running it live would error.
See the [Factor Models & BHAR](../articles/factor-models-bhar.html) vignette for
a worked factor example.

```{r ff-carhart, eval=FALSE}
# Requires a factor_tbl with market_excess/smb/hml[/rmw/cma/mom]:
params_ff3 <- ParameterSet$new(return_model = FamaFrench3FactorModel$new())
params_ff5 <- ParameterSet$new(return_model = FamaFrench5FactorModel$new())
params_c4  <- ParameterSet$new(return_model = Carhart4FactorModel$new())
```

The factor models decompose systematic exposure beyond a single market beta
[@FamaFrench1993; @FamaFrench2015; @Carhart1997], reducing the risk that a
size/value/momentum tilt is mislabelled as abnormal performance.

**GARCH** conditions the residual variance on its own recent history, which
matters when volatility clusters around the event. It is gated on `rugarch`:

```{r garch, eval=requireNamespace("rugarch", quietly=TRUE)}
params_g <- ParameterSet$new(return_model = GARCHModel$new())
task_g <- prepare_event_study(
  EventStudyTask$new(dieselgate$firm, dieselgate$index, dieselgate$request),
  params_g
)
task_g <- fit_model(task_g, params_g)
task_g <- calculate_statistics(task_g, params_g)
es_tt(head(tidy.EventStudyTask(task_g, type = "car"), 5),
      caption = "GARCH-based CAR t-statistics.")
```

If `rugarch` is not installed the chunk above is skipped. Conceptually, the
GARCH model replaces the constant OLS residual variance $\sigma^2$ with a
time-varying $\sigma_t^2$ following a GARCH(1,1) recursion, and the
**DCC-GARCH** extension further models the time-varying correlation across
firms — appropriate when a single event hits several securities whose joint
volatility co-moves.

**Volume and Volatility** models target abnormal trading volume and abnormal
return dispersion rather than abnormal price returns; see the
[Volume & Volatility Event Study](../articles/volume-volatility-event-study.html)
vignette for worked usage.

```{r vol-conceptual, eval=FALSE}
# Volume / Volatility models need firm_volume columns not present in dieselgate:
params_vol <- ParameterSet$new(return_model = VolumeModel$new())
```

## 9. Interpretation

The table in §7 reports the per-horizon CAR and its $t$-statistic for the
dieselgate event; a $t$ beyond roughly $\pm 2$ flags a statistically abnormal
cumulative return at conventional levels. The plot in §8 traces the CAR path
across the event window — a sharp downward break at the event date is the
visual signature of the emissions-scandal shock. Choose the factor or GARCH
variants when the market-model assumptions in §5 are implausible for your
setting; the abnormal-return machinery downstream is identical.

## References
