---
title: "Methods: AI Advisor"
output: rmarkdown::html_vignette
bibliography: references.bib
# pkgdown article — excluded from the tarball via .Rbuildignore (^vignettes/articles).
---

```{r child="_setup.Rmd"}
```

```{r advisor-offline, include=FALSE}
# Keep this article offline and API-key-free at build time: the LLM layer is
# eval=FALSE, and the deterministic layer never touches the network.
Sys.setenv(EVENTSTUDY_NO_NETWORK = "1")
```

## 1. Title & Abstract

This article describes the **two-layer AI advisor**: a deterministic,
fully-offline grounding layer that computes diagnostics and rule-based
recommendations from the package's own numbers, and an optional LLM layer that
narrates — but never invents — those numbers. After reading, you will
understand the advisor's design and the grounding invariant that keeps it
trustworthy, and you will see the deterministic layer run live on `dieselgate`.

## 2. When to Use This Method

Use the advisor when you want a **defensible interpretation** of an event study
without hand-writing the diagnostic checklist: which test statistic is
appropriate given the normality, autocorrelation, overlap, and cross-sectional
spread of your data, and where the robustness caveats lie. The deterministic
layer alone is enough for reproducible, offline advice; the LLM layer only adds
prose.

## 3. Intuition

Two layers, one direction of trust. The deterministic layer measures the study
(Shapiro-Wilk normality, Ljung-Box autocorrelation, $R^2$, CAR dispersion,
window overlap) and maps those measurements to recommendations through fixed
rules. The LLM layer is handed **only** those computed facts and asked to
phrase them; it is structurally prevented from producing a number the
deterministic layer did not.

## 4. Model & Null Hypothesis (Grounding Invariant)

This is a **design** article, not a statistical estimator, so it carries no
academic formula. Its central claim is the advisor's grounding invariant,
quoted from the `ai-advisor` vignette:

> The advisor never fabricates a number. Every claim it returns is provably tied
> to a package-computed diagnostic.

The invariant is enforced by a runtime guard in R, independent of the LLM: the
recommendation and robustness flags derive deterministically from
`es_diagnostics()`, so they are identical whether or not any model is
contacted. There is no null hypothesis to test here — the "null" the design
defends against is a hallucinated statistic, and the guard makes that
impossible by construction.

## 5. Assumptions

- **Fitted pipeline:** the task has been through
  `fit_model()` / `calculate_statistics()` before advising.
- **Offline determinism:** with `EVENTSTUDY_NO_NETWORK=1`, the deterministic
  layer is a pure function of the task — reproducible across runs and machines.
- **LLM layer is narration-only:** it receives grounded facts, never raw data,
  and cannot introduce new numbers.

## 6. Worked Example

The deterministic layer runs fully offline on `dieselgate`.

```{r load-data}
library(EventStudy)
data("dieselgate")
```

```{r fit-pipeline}
task   <- EventStudyTask$new(dieselgate$firm, dieselgate$index, dieselgate$request)
params <- ParameterSet$new()
task   <- prepare_event_study(task, params)
task   <- fit_model(task, params)
task   <- calculate_statistics(task, params)
```

```{r deterministic-layer}
diag   <- es_diagnostics(task)     # six-section diagnostics list, offline
advice <- recommend_stat(task)     # rule-based test recommendation
robust <- flag_robustness(task)    # robustness caveats
```

The optional **LLM layer** mirrors `vignettes/ai-advisor.Rmd`: it is
`eval=FALSE` and shown with static captured output so the build never contacts
the network.

````{r llm-layer, eval=FALSE}
# Requires an API key + network; disabled at build time.
narrative <- es_advise(diag, task_type = "interpret", provider = provider("anthropic"))
narrative
#> The market-model fit is strong (median R^2 ~ 0.6). Residual normality is not
#> rejected for most firms, so the parametric CAR t-test is appropriate; the
#> Patell Z is offered as a standardized cross-check. No estimation-window
#> overlap was detected, so cross-sectional correlation is a minor concern.
````

## 7. Rendered Table

The `estimation_window` section of the diagnostics — the per-event fit and
residual checks the advisor reasons over:

```{r results-table}
est <- as.data.frame(diag$estimation_window)
es_tt(
  est,
  caption = "Per-event estimation-window diagnostics feeding the advisor (offline)."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Diagnostic panel used by the deterministic advisor layer."}
plot_diagnostics(task)
```

## 9. Interpretation

The table in §7 is the ground truth the advisor stands on: each row's $R^2$,
residual sigma, Shapiro-Wilk $p$, Durbin-Watson statistic, and Ljung-Box $p$
determine which test the rule layer recommends and which caveats it raises.
Because `advice` and `robust` are pure functions of these numbers, re-running
the chunk reproduces them exactly — and the LLM layer, if enabled, can only
restate them. That is the grounding invariant in action: the interpretation is
as trustworthy as the diagnostics, and never more.

## References
