---
title: "Methods: Diagnostics & Robustness"
output: rmarkdown::html_vignette
bibliography: references.bib
# pkgdown article — excluded from the tarball via .Rbuildignore (^vignettes/articles).
---

```{r child="_setup.Rmd"}
```

## 1. Title & Abstract

This article covers the **diagnostics and robustness** tools that validate an
event study's inferential assumptions: residual normality (Shapiro-Wilk),
autocorrelation (Durbin-Watson and Ljung-Box), pre-event trends, wild-bootstrap
inference, and Monte-Carlo power analysis. After reading, you will know how to
check whether the parametric $t$-tests are trustworthy for your data and how to
fall back on bootstrap inference when they are not. All examples run live and
offline on `earnings_surprises` and simulated data.

## 2. When to Use This Method

Run diagnostics **before** trusting any test statistic. Normality and
autocorrelation checks tell you whether the parametric CAR $t$-test's
distributional assumptions hold; a pre-trend test guards against a
contaminated estimation window. When assumptions fail, the **wild bootstrap**
provides distribution-free $p$-values, and a **power simulation** tells you
whether your design could have detected the effect you care about.

## 3. Intuition

The abnormal-return machinery assumes clean, roughly-normal, serially
uncorrelated residuals. Diagnostics interrogate each of those assumptions. If
they hold, the closed-form $t$-tests are fine. If they don't, resample: the
wild bootstrap perturbs the observed abnormal returns with random $\pm 1$
weights to build an empirical null, sidestepping the normality assumption
entirely.

## 4. Model & Null Hypothesis

**Ljung-Box** tests residual autocorrelation up to lag $h$:

<!-- Formula verified: R/diagnostics.R:64-67 (Box.test type=Ljung-Box) matches BoxLjung1978 -->

$$
Q = n(n+2) \sum_{k=1}^{h} \frac{\hat{\rho}_k^2}{n-k}.
$$

**Durbin-Watson** targets first-order serial correlation of the residuals
$e_t$ (the package computes the approximate statistic directly):

<!-- Formula verified: R/diagnostics.R:56-61 matches DurbinWatson1950 -->

$$
d = \frac{\sum_{t=2}^{n} (e_t - e_{t-1})^2}{\sum_{t=1}^{n} e_t^2}.
$$

The **wild bootstrap** resamples abnormal returns with Rademacher weights
$w_b \in \{-1, +1\}$:

<!-- Formula verified: R/bootstrap.R:91-93 matches (wild bootstrap) -->

$$
AR^{*}_{b} = w_b\, \hat{AR}, \qquad w_b \sim \text{Rademacher}.
$$

**Power** is the Monte-Carlo rejection rate under a specified true effect:

<!-- Formula verified: R/simulation.R:115 matches (Monte Carlo power) -->

$$
\text{power} = \frac{1}{S}\sum_{s=1}^{S} \mathbf{1}\{\text{reject } H_0 \text{ in sim } s\}.
$$

The null hypotheses are: $H_0: \hat{\rho}_k = 0$ (no autocorrelation,
Ljung-Box / DW $d \approx 2$) and $H_0: \mathbb{E}[AR] = 0$ (no abnormal return,
bootstrap and power) [@BoxLjung1978; @DurbinWatson1950].

## 5. Assumptions

- **Diagnostics are per-event:** normality and autocorrelation are checked on
  each event's estimation-window residuals.
- **DW $\approx 2$** indicates no first-order autocorrelation; values far from 2
  flag it.
- **Wild bootstrap** relaxes normality but still assumes the abnormal returns
  are the correct point estimates to resample.
- **Power simulation** is only as informative as its assumed effect size and
  noise structure.

## 6. Worked Example

Diagnostics run live on `earnings_surprises`; bootstrap and power run with an
explicit `seed = 42` for byte-stability.

```{r load-data}
library(EventStudy)
data("earnings_surprises")
```

```{r fit-pipeline}
task   <- EventStudyTask$new(
  earnings_surprises$firm, earnings_surprises$index, earnings_surprises$request
)
params <- ParameterSet$new()
task   <- prepare_event_study(task, params)
task   <- fit_model(task, params)
task   <- calculate_statistics(task, params)
```

```{r diagnostics}
diag <- model_diagnostics(task)   # Shapiro-Wilk, Durbin-Watson, Ljung-Box per event
pt   <- pretrend_test(task)       # pre-event trend test
```

```{r robustness}
# Power on a simulated design (zero-network, seed-stable):
sim  <- simulate_event_study(n_events = 20, abnormal_return = 0.02,
                             n_simulations = 200, seed = 42)

# Wild bootstrap inference on the observed task:
boot <- bootstrap_test(task, n_boot = 199, seed = 42)
```

## 7. Rendered Table

```{r results-table}
es_tt(
  diag,
  caption = "Per-event diagnostics: Shapiro-Wilk (shapiro_p), Durbin-Watson (dw_stat), Ljung-Box (ljung_box_p)."
)
```

## 8. Rendered Plot

```{r results-plot, fig.cap="Monte-Carlo rejection rate (power) by event-window day."}
plot(sim$rejection_by_day$relative_index,
     sim$rejection_by_day$rejection_rate,
     type = "b", pch = 19,
     xlab = "Relative day", ylab = "Rejection rate",
     main = "Simulated power by event-window day")
abline(h = 0.05, lty = 2, col = "grey60")
```

## 9. Interpretation

Read §7 row by row: a small `shapiro_p` rejects residual normality (prefer the
bootstrap or a nonparametric test); a `dw_stat` far from 2 or a small
`ljung_box_p` flags autocorrelation (inference standard errors are then
understated). The power curve in §8 shows how the design's detection rate rises
around the event day for the assumed effect — a curve that never lifts much
above the dashed 0.05 line means the study is underpowered for that effect
size. When the diagnostics fail, the wild-bootstrap $p$-values in `boot`
(`boot_p_aar`, `boot_p_caar`) are the assumption-light fallback for reporting
significance.

## References
