---
title: "Methods Gate: Smoke Test"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Methods Gate: Smoke Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="_setup.Rmd"}
```

## Cumulative Abnormal Return Formula

The Cumulative Abnormal Return (CAR) over the event window $[t_1, t_2]$ is defined as:

$$\text{CAR}(t_1, t_2) = \sum_{t=t_1}^{t_2} \hat{\varepsilon}_t$$

where $\hat{\varepsilon}_t = R_{it} - (\hat{\alpha} + \hat{\beta} R_{mt})$ is the abnormal return
under the market model [@MacKinlay1997].

The market model regresses firm returns on an index during an estimation window, then uses
the out-of-sample residuals in the event window as abnormal returns [@Brown1985].

## Worked Example: Dieselgate

This example uses the bundled `dieselgate` dataset (4 automakers, 2 groups) to demonstrate
the full EventStudy pipeline.

```{r load-data}
library(EventStudy)
data("dieselgate")
```

```{r run-pipeline}
task <- EventStudyTask$new(
  firm_stock_data_tbl = dieselgate$firm,
  reference_tbl       = dieselgate$index,
  request_tbl         = dieselgate$request
)
params <- ParameterSet$new()
task   <- prepare_event_study(task, params)
task   <- fit_model(task, params)
task   <- calculate_statistics(task, params)
```

The pipeline fits a market model for each firm against the DAX (^GDAXI) using the
estimation window, then computes CAR t-statistics for the event window around
18 September 2015 (VW emissions revelation).

```{r plot-car, fig.cap="CAR over event window (dieselgate, event 1)"}
car_plot <- plot_event_study(task, type = "car", event_id = 1)
plotly::ggplotly(car_plot)
```

## References
